Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-04-21 manuscript/manuscript.tex
updated figures, ms

In analyzing both the empirical datasets included in our study and our
simulation results, we noticed the temporal covariances and $G(t)$ statistics
are sensitive to how fixations are handled. Through simulations and comparing
the results of analyzing our empirical datasets both ways, we have found
tradeoffs to both approaches, but overall not removing fixed sites is a more
conservative, safer approach, and is what we use throughout our paper.

Temporal covariances (and related statistics like convergence correlations,
$G(t)$, etc.) are calculated on allele frequency changes, $\Delta p_t$; when an
allele is fixed or lost, the allele frequency changes are zero thereafter
$\Delta p_t = 0$. When temporal covariance is calculated by averaging over the
loci in a region or across the genome, fixed sites zero out the terms in the
sum of the right hand side of temporal covariance, $\cov(\Delta p_t, \Delta
p_s) = \E(\Delta p_t \Delta p_s) - \E(\Delta p_t) \E(\Delta p_s)$ and can
shrink the second terms towards zero.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth]{figures/fig-bgs-covars-with-fixations.pdf}

>>>>

In analyzing both the empirical datasets included in our study and our
simulation results, we noticed the temporal covariances and $G(t)$ statistics
are sensitive to how fixations (i.e. when a site is polymorphic but is fixed or
lost during the course of its trajectory) are handled. Through simulations and
comparing the results of analyzing our empirical datasets both ways, we have
found trade-offs to both approaches, but overall find that not removing fixed
sites is a more conservative, safer approach, and is what we use throughout our
paper.

@@

  \caption{$G(t)$ trajectories from background selection simulations both
  including fixed sites (solid lines) and not including fixed sites (dashed
lines). The dashed lines are identical to those in Supplementary Materials
Figure \ref{suppfig:supp-bgs-g}. As expected, including fixed sites shrinks the
$G(t)$ towards zero, and in some cases, they can become negative.}

  \label{suppfig:supp-bgs-g-fix}
\end{figure}

Temporal covariances (and related statistics like convergence correlations,
$G(t)$, etc.) are calculated on allele frequency changes, $\Delta p_t$; when an
allele is fixed or lost, the allele frequency changes are zero thereafter
$\Delta p_t = 0$. When temporal covariance is calculated by averaging over the
loci in a region or across the genome, fixed sites zero out the terms in the
sum of the right hand side of temporal covariance, $\cov(\Delta p_t, \Delta
p_s) = \E(\Delta p_t \Delta p_s) - \E(\Delta p_t) \E(\Delta p_s)$ and can
shrink the second terms towards zero. From simulations using population
frequencies (e.g. no sampling noise), we have found that including neutral
sites into the temporal covariance calculation that are lost or fix will tend
to shrink the covariance towards zero, and in some cases create spurious
negative temporal covariances. We observed this when conducting the background
selection simulations of Supplementary Materials Section XXX, and have included
an alternative version of this figure that includes $G(t)$ trajectories
calculated by including fixed neutral sites (Supplementary Figure
\ref{suppfig:supp-bgs-g-fix}). We find, as expected, $G(t)$ trajectories
calculated including fixed sites are shrunk towards zero, and in some cases,
can become negative. In contrast, the $G(t)$ trajectories ignoring fixations
(dashed lines in Supplementary Figure \ref{suppfig:supp-bgs-g-fix}) are higher.

XXX truncation selection, exponential.

While simulation results using population frequencies show that including fixed
sites has the downside of shrinking the temporal covariance towards zero, and
in some cases can create negative covariances, we have found excluding fixed
sites on sample frequency data has a severe drawback: low frequency alleles
will occasionally not be observed at a sample timepoint (leading to a frequency
of zero of the minor allele), and excluding these observations (instead of
treating it as a trajectory that has a 0 frequency timepoint) biases estimates.
We observed this bias created by ignoring fixed or loss sites by comparing our
$N_e$ estimates for the \textcite{Barghi2019-qy} study (estimated from the
total variance in allele frequency change) with the estimates from the original
paper. We found that ignoring fixed or lost sites (by treating them as missing
data in calculating the covariances) led to many low-frequency alleles not
contributing to the variance, leading this statistic to be calculated on more
intermediate frequency alleles and thus biasing the $N_e$ estimate downwards.
Additionally, we tried only dropping fixed or loss sites from the temporal
covariance calculations that were at the end or the beginning of a trajectory
(e.g., as if the site was created by a new mutation or fixed); while this
ameliorated some of this bias, it still did not lead to $N_e$ estimates
congruent with the original studies. Overall, we found by trying all these
approaches that not removing fixed or lost sites was the best way to deal with
sample allele frequencies that could be missing from some timepoints.

Still, we wanted to ensure that our empirical re-analyses were robust to our
choice of how we handled fixed sites. Given that under some simulation
conditions, we find that spurious negative covariances can occur when fixations
are included, we wanted to ensure that our result that the temporal covariances
of \textcite{Barghi2019-qy} that become negative at later timepoints are not an
artifact of the way fixations are handled. First, we regenerated Figure 1 A and
B, except with temporal covariances and $G(t)$ calculated excluding fixed or
lost sites. We find the same qualitative results regardless of whether we
include or exclude fixed/lost sites; importantly, we still see a significant
signal that the temporal covariances at later timepoints become negative.

XXX

\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.6\textwidth]{figures/figure-1-G-covs-nofix.pdf}

  \caption{A version of Figure 1 A and B where sample frequencies of zero or
    one are ignored, e.g. due to fixation, are ignored. Like the original
    Figure 1 in the main text, the 95\% pivot confidence intervals are via a
    bootstrapping procedure; note that we find excluding fixed sites leads to
  more asymmetric confidence intervals. Overall, we find that despite handling
fixed/lost sites differently, we still find significant negative temporal
covariances at later timepoints.}

  \label{suppfig:supp-fig-1-nofix}
\end{figure}
